\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2018-01-19 eg6-letters/NIM/EG6_RTPC_NIM.tex
Last (?) edits on the NIM paper

The timing information collected for each signal above threshold
was used to infer the origin of the ionization electrons and 
then the trajectory of the detected particle. The recorded ADCs were used to reconstruct 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enabled the particle 
identification.

In this section we will detail the methods used to calibrate the drift time,
drift paths and gains of the detector. Drift time and paths were initially
calculated using the MAGBOLTZ~\cite{Biagi:1999nwa} program, then refined using
data to account for variations of the run conditions. We assume 
cylindrical symmetry in the chamber for the calibration of the drift parameters, 
such that do not depend on the azimuthal angle $\phi$. The initial MAGBOLTZ
calibration was improved through several iterations of the
data driven process described below, with each time an increasing number of tracks 
reconstructed in the RTPC. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process.

@@

%To be rewritten!!!
After the ionization, the released electrons drift to the cylindrical detection 
plane under the effect of the electric field. The electrons released close to 
the cathode take the most time to reach the readout pads, the cylindrical 
symmetry indicates that this maximum drift distance is the same for all tracks.  
We illustrate this in Figure~\ref{fig:RTPC_signals}, where we depict a typical 
$^{4}$He track. Due to the non-perfect experimental conditions, in particular 
possible contamination of our gas mixture, the drift speed changed during the 
three months long experimental run\footnote{We had to increase the gas flow 
during the experiment due to a small leak in the RTPC, which concurred to 
the large shift of drift time observed around run number 61600. While our 
gas system was kept slightly over atmospheric pressure to limit 
contamination from air or other external gases, it is likely that this leak was 
the source of modification of the drift speed.}. This can be calibrated by 
finding drift speed as function of experimental run-number. By measuring a 
reference time, we can therefore infer the drift speed of the electrons in the 
RTPC, by: $\textnormal{drift time} = \frac{\textnormal{drift path 
length}}{\textnormal{drift speed}}$.

@@

Figure~\ref{fig:TDC_profile} presents the time profile of hits from identified 
tracks in one z-bin along the detector in one experimental run. We can clearly 
observe the dropping edge expected from geometrical considerations. In order to 
avoid the statistical fluctuations on finding the maximum drift time 
($T_{max}$) from the time profile, we define a more stable value, i.e., 
$T_{max/2}$, at which the dropping edge passes half the maximum number of hits 
in the histogram. This value was measured in bins along the 200~mm RTPC's 
length to take into account variations in the electric and magnetic field in 
the RTPC (see Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}).

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption[]{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. Trigger time is defined as T=15, the time unit is the ALTRO 
sampling time of 100~ns.} \label{fig:TDC_profile}
\end{figure}

@@

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max/2}$ values for individual 
runs (approximately 2 hours long). We observe significant change in the drift 
speed before and after run 61600, while variations within these periods are 
around 2\%.

@@

In summary, we obtain from our calibration a parametrization of the drift speed 
as a function of both position along the beam axis and the run number.  These 
functions were extracted for our entire data set and implemented in the track 
reconstruction code.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We calculated them with
MAGBOLTZ~\cite{Biagi:1999nwa}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used this calculation as a 
first calibration, but, as can be seen with the drift speed, conditions in the 
chamber were changing over time. Moreover, the 4 \textmu{}m foil used as a cathode 
is easily deformed, such that we expect the geometrical accuracy to be of few 
millimeters, directly impacting our knowledge of the electric field. These 
problems, already encountered for the BoNuS RTPC 
calibration~\cite{Fenker:2008zz}, motivated the acquisition of specific 
calibration runs. These were taken with a lower energy electron beam (1.204 and 
1.269 GeV) to enhance the cross section of the elastic scattering 
($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate directly the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

Based on the kinematics of the electrons in the calibration data, we generated 
the Helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then,
we compared the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. To perform this drift path extraction, 
we made a first approximation assuming a linear dependence between the radius 
of emission of the charge and its time of detection, and then refined our 
result. Because of the magnetic field, the drift paths in RTPC are not linear 
but the curvature is small and the refining  process converged already on the 
second iteration.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Figure~\ref{fig:Drift_run_number_1}). We found very similar drift paths for the 
two data sets and concluded that any changes in the system only significantly 
affected the drift speed.

@@

To calibrate the gains, we compared the experimental ADCs to the energy 
deposited for each pad individually in GEANT4 by similar simulated tracks 
(using the same elastic events as for the drift paths calibration). This 
requires a detailed simulation, including the drift paths and the diffusion of 
the charges along the path before reaching the pad to 
match the experimental data. We implemented in the GEANT4 simulation the
drift path we obtained from the calibration described above and implemented an ad 
hoc diffusion to match the average number of hits recorded in the simulation with
the one from experimental data. To perform this step properly, we implemented in
the simulation the details of the DAQ process to record hits.

We then compared this simulation to experiment on an event by event basis as 
shown in Figure~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy. This 
provided our initial set of gains for each pads and it was applied to the 
experimental data. However, after this calibration some pads recorded lower ADCs 
than expected. So, we added a correction factor obtained by comparing  
the energy deposit on a given pad to its neighbours. To do so, we compared, for 
each track, the energy deposit on a pad to the energy deposit in the whole track. 
The gain correction factors are extracted by averaging this ratio for a sample 
of elastic events. Final results are shown in Figure~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see the band for $^4$He in its expected position\footnote{The lower peak 
   in the left module of the RTPC comes from an unidentified problem in this half 
   of the detector that concerns 7\% of all the elastic events. 
   Our best guess is that these events are linked with a high voltage supply 
   issue, that would sometimes bring the GEM gain down. These events pass all 
   the elastic requirements and we found nothing that differenciates them 
   from other events but their low recorded ADCs. For 
   the calibration procedures, these events were therefore excluded.}.

>>>>

The timing information collected from each signal is used 
to infer the origin of the ionization electrons, from which we
reconstruct the trajectory of the initial particle. The strength of these 
signals, recorded in ADC units, is then used to reconstruct 
the deposited energy per unit of length ($\small{\frac{dE}{dX}}$) which, 
together with the momentum calculated from the trajectory, enabled the particle 
identification.

In this section we will detail the methods used to calibrate the drift time,
drift paths and gains of the detector. The drift paths were initially
calculated using the MAGBOLTZ~\cite{Biagi:1999nwa} program, then refined using
data to account for variations of the run conditions. The initial MAGBOLTZ
calibration was improved through several iterations of the
data driven process described below, with each time an increasing number of tracks 
reconstructed in the RTPC justifying a new iteration. The figures presented in this section
are the ones obtained while performing the last iteration of this
calibration process. We assume 
cylindrical symmetry in the chamber for the calibration of the drift parameters, 
such that do not depend on the azimuthal angle $\phi$.

@@

The first step in the calibration is to fix the maximum drift time ($T_{max}$) that an
electron can arrive on the detection plane. This value is highly dependent on the gas 
mixture composition, the electric and the magnetic field experienced by the drift 
electrons. This time correspond to the electrons released close to 
the cathode, which take the most time to reach the readout pads as illustrated in 
Figure~\ref{fig:RTPC_signals}. Note that the minimum time is not calibrated, as we
use the trigger time is defined ($T_{min}=15$), and that the time unit is the ALTRO 
sampling time of 100~ns.

@@

In order to experimentally determine the value of $T_{max}$, we analyse the
time profile of hits from identified good tracks, as presented in 
Figure~\ref{fig:TDC_profile}. We can clearly observe the expected dropping 
edge and define $T_{max}$, as the time at which the distribution passes 
below half the maximum number of hits in the histogram. This value was 
measured in bins along the 200~mm RTPC's 
length to take into account variations in the electric and magnetic field in 
the RTPC. We show this result for one experimental run in 
Figure~\ref{fig:RunNumber_61551_TDCmax_Zslice}.

\begin{figure}[tb]
\centering
\includegraphics[scale=0.42]{hits_time_profile.pdf}
\caption{Time distribution of the hits in one z-bin along the RTPC in one 
experimental run. } \label{fig:TDC_profile}
\end{figure}

@@

Figure~\ref{fig:Drift_run_number_1} shows the $T_{max/2}$ values for individual 
runs (approximately 2 hours long). We observe significant change of $T_{max}$
before and after run 61600, while variations within these periods are 
around 2\%. This is likely due to non-perfect experimental conditions, in 
particular possible contamination of our gas mixture. We indeed increase the gas flow 
around that time due to a small leak in the RTPC. While our 
gas system was kept slightly over atmospheric pressure to limit 
contamination from air or other external gases, it is likely that this leak was 
the source of modification of the drift velocity.

@@

To take this effects into account, we parametrized the maximum drift time 
as a function of both the position along the beam axis and the run number. These 
functions were extracted for the entire data set and implemented in the track 
reconstruction code for the next steps of calibration and track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. We initially calculated them with
MAGBOLTZ~\cite{Biagi:1999nwa}, but this calculation requires knowledge of the detector's 
geometry, of the gas mixture composition, and of the electric and magnetic 
fields over the whole volume of the detector. As we seen in the previous section, 
the conditions in the 
chamber were changing over time. Moreover, the 4 \textmu{}m foil used as a cathode 
is easily deformed, such that we expect the geometrical accuracy to be only of a few 
millimeters, also impacting our knowledge of the electric field. These 
problems, already encountered for the BoNuS RTPC 
calibration~\cite{Fenker:2008zz}, motivated the acquisition of specific 
calibration runs. These were taken with a lower energy electron beam (1.204 and 
1.269 GeV) to enhance the cross section of the elastic scattering 
($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematics allows to calculate the kinematics of the Helium nucleus. 
By comparing the calculated momentum and angle of the recoil alpha particle to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the chamber's conditions.

Based on the kinematics of the electrons in the calibration data, we generated 
the Helium nucleus in our RTPC GEANT4 simulation~\cite{GEANT4}. Then,
we compared the calculated GEANT4 trajectory of the Helium nuclei to 
the hits measured in the chamber. To perform the drift path extraction, 
we made a first approximation assuming a linear dependence between the radius 
of emission of the charge and its time of detection, and then refined our 
result. Because of the magnetic field, the drift paths in the RTPC are not linear 
but the curvature is small and the process converged already on the 
second iteration.

@@

To verify the stability of the drift paths, this procedure was carried out 
using both the 1.204 GeV data from the beginning of the run period and the 
1.269 GeV data from the end of the run period (shown in blue on 
Figure~\ref{fig:Drift_run_number_1}). We found very similar drift paths for the 
two data sets and concluded that any changes in the system only significantly 
affected the drift speed and thus the measured maximum time.

@@

To calibrate the gains, we compared the experimental ADCs to the energy 
deposited for each pad individually in GEANT4 by similar simulated tracks 
(using the same elastic events as for the drift paths calibration). This 
requires a detailed simulation, including the drift paths and the diffusion of 
the charges along the path before reaching the pad to 
match the experimental data. We implemented in the GEANT4 simulation the
drift path we obtained from the calibration described above and implemented an {\it ad 
hoc} diffusion function to match the average number of hits recorded in the simulation with
the one from experimental data. To perform this step properly, we also implemented in
the simulation the details of the DAQ process to record hits.

We then compared this simulation to experiment on an event by event basis as 
shown in Figure~\ref{fig:EVENT_adc_tdc}. The gain for each pad was calculated 
as the ratio of the measured ADCs to the simulated deposited energy. This 
provided our initial set of gains for each pads and it was applied to the 
experimental data. However, after this calibration some pads recorded lower ADCs 
than expected. So, we added a correction factor obtained by comparing  
the energy deposit on the pads within individual tracks. To do so, we compared, for 
each track, the energy deposit on a pad to the energy deposit in the whole track. 
The gain correction factors are extracted by averaging this ratio for a sample 
of elastic events and are thus completely independent from simulation.

The results after calibration are shown in Figure~\ref{fig:dedx_p_exp_2nd}, where energy 
loss of particles is plotted against momentum over charge ratio. One can 
clearly see there the band for $^4$He in its expected position\footnote{The lower peak 
   in the left module of the RTPC comes from an unidentified problem in this half 
   of the detector that concerns 7\% of all the elastic events. 
   Our best guess is that these events are linked with a high voltage supply 
   issue, that would sometimes bring the GEM gain down. These events pass all 
   the elastic requirements and we found nothing that differenciates them 
   from other events but their low recorded ADCs. In particular, 94\% of pads
   are involved with these tracks such that we are sure they do not come from 
   a miscalibration of a part of the detector. Finally, we decided to discard
   these events for the calibration procedures.}.
